---
title: "Getting started with claimtestR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with claimtestR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`claimtestR` represents a scientific claim as an executable assertion. Unlike a
testthat expectation, an ordinary claim does not stop execution when it fails.
It returns evidence that can be inspected, summarized, and included in a report.

## Fit a model

This vignette uses a base R linear model so that the full workflow is
reproducible without downloading data or installing modeling packages.

```{r model}
library(claimtestR)

model <- lm(mpg ~ wt + hp, data = mtcars)
summary(model)$coefficients
```

## Declare claims

Each assertion identifies the model term explicitly. The practical-effect claim
below requires a decrease of at least two miles per gallon per unit of weight.

```{r claims}
claims <- test_claims(
  expect_negative_effect(model, term = "wt"),
  expect_interval_excludes(model, term = "wt", value = 0),
  expect_practical_effect(
    model,
    term = "wt",
    minimum = 2,
    direction = "negative"
  )
)

claims
```

Every result contains a stable claim identifier, status, observed value,
expectation, term, message, evidence, diagnostic details, and originating call.

```{r inspect}
str(claims$results[[1]], max.level = 1)
summary(claims)
```

## Descriptive versus inferential claims

`expect_negative_effect()` and `expect_practical_effect()` describe the point
estimate only. Interval-based claims state the same conclusions with
uncertainty taken into account. Here a smaller `wt` coefficient is "better"
(a stronger decrease in fuel economy per unit of weight), so the claims use
`direction = "lower"`. The margin and the confidence level are chosen by the
analyst: a 90% interval corresponds to one-sided tests at 0.05.

```{r inferential}
inferential <- test_claims(
  expect_superior(model, term = "wt", direction = "lower", level = 0.90),
  expect_superior(model, term = "wt", margin = 2, direction = "lower",
                  level = 0.90),
  expect_noninferior(model, term = "wt", margin = 1, direction = "lower",
                     level = 0.90)
)

inferential
```

Both superiority claims pass because the whole 90% interval, from about
-4.95 to -2.80, lies below zero and also below -2. The second claim is
stricter than `expect_practical_effect()` with the same threshold, which only
compares the point estimate with -2.

## Report and export results

`report_claims()` returns a Markdown table suitable for a Quarto or R Markdown
document.

```{r markdown-report, results = "asis"}
cat(report_claims(claims, format = "markdown"), "\n")
```

For further formatting or archival export, convert the suite to a data frame.

```{r export}
claim_data <- as.data.frame(claims)
claim_data

export_path <- tempfile(fileext = ".csv")
utils::write.csv(claim_data, export_path, row.names = FALSE)
file.exists(export_path)
unlink(export_path)
```

The data frame can also be passed directly to `knitr::kable()`.

```{r kable, results = "asis"}
knitr::kable(claim_data)
```

## Make a claim enforceable in CI

Use `expect_claim_passes()` inside a testthat test when a failed scientific claim
must fail continuous integration.

```{r testthat-bridge}
if (requireNamespace("testthat", quietly = TRUE)) {
  expect_claim_passes(
    expect_negative_effect(model, term = "wt")
  )
}
```

For example, a project can place the following code in `tests/testthat/`:

```r
testthat::test_that("the primary conclusion remains valid", {
  model <- run_analysis()
  claimtestR::expect_claim_passes(
    claimtestR::expect_negative_effect(model, term = "treatment")
  )
})
```

## Check stability

The initial stability API accepts existing resample estimates or can refit base R
models with row bootstrap and leave-one-out analyses.

```{r stability}
stability <- expect_stable_direction(
  model,
  term = "wt",
  method = "bootstrap",
  iterations = 50,
  minimum_proportion = 0.9,
  seed = 2026
)
stability
```

Model-based stability requires at least two successful refits and a success
proportion of at least 80%. Failed refits are not discarded from the stability
denominator. If the threshold is not met, the function returns a failed
`claim_test` and explains that the assessment is unreliable. The diagnostic
counts and captured conditions remain available for auditing:

```{r stability-diagnostics}
stability$details[c(
  "attempted", "successful", "failed", "success_proportion",
  "minimum_success_proportion", "resampling_valid"
)]
stability$details$failures
stability$details$warnings
```

Refits use the original variables after aligning them to the observations that
the fitted model retained. Transformations are therefore reevaluated inside each
resample rather than treated as precomputed columns:

```{r transformed-stability}
transformed_model <- lm(mpg ~ log(wt), data = mtcars)
transformed_stability <- expect_stable_direction(
  transformed_model,
  term = "log(wt)",
  method = "bootstrap",
  iterations = 20,
  minimum_proportion = 0.8,
  seed = 2026
)
transformed_stability
```

For stability refits in version 0.1.2, `data` must be a simple name that points
to an ordinary base data frame whose columns are unclassed atomic vectors or
base factors. Active bindings, additional data or column classes, and all calls
in `data` -- including `subset()`, `transform()`, `within()`, and `get()` -- are
rejected rather than evaluated. Materialize the result first:

```{r materialized-data}
filtered_data <- subset(mtcars, mpg > 15)
filtered_model <- lm(mpg ~ log(wt), data = filtered_data)
expect_stable_direction(
  filtered_model,
  term = "log(wt)",
  method = "bootstrap",
  iterations = 20,
  seed = 2026
)
```

Formula objects stored under local names remain supported: claimtestR uses the
formula object retained by the fitted model, including its environment, instead
of depending on the original formula symbol still being available.

Models fit with `model = FALSE` are not eligible for refitting because they do
not retain the model frame needed to identify the observations actually used.

Bootstrap refits of a `glm` that do not converge are recorded as failures. A
non-converged original `glm` is rejected before any claim is evaluated;
claimtestR does not repair or reinterpret that model. When `seed` is supplied,
the bootstrap result is reproducible and the caller's random-number state is
restored even if evaluation exits with an error.

Row bootstrap assumes that rows are the appropriate independent sampling units.
Clustered and longitudinal analyses require cluster-aware resampling, which is
planned for a later release and must not be approximated with this method.

## Scope and limitations

`claimtestR` checks claims chosen by the analyst. It does not infer claims from
prose, select a statistical model, determine whether observations are
independent, or establish that a chosen threshold is scientifically justified.
Those decisions remain part of the study design and analysis plan.
