---
title: "Getting started with claimtestR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with claimtestR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`claimtestR` represents a scientific claim as an executable assertion. Unlike a
testthat expectation, an ordinary claim does not stop execution when it fails.
It returns evidence that can be inspected, summarized, and included in a report.

## Fit a model

This vignette uses a base R linear model so that the full workflow is
reproducible without downloading data or installing modeling packages.

```{r model}
library(claimtestR)

model <- lm(mpg ~ wt + hp, data = mtcars)
summary(model)$coefficients
```

## Declare claims

Each assertion identifies the model term explicitly. The practical-effect claim
below requires a decrease of at least two miles per gallon per unit of weight.

```{r claims}
claims <- test_claims(
  expect_negative_effect(model, term = "wt"),
  expect_interval_excludes(model, term = "wt", value = 0),
  expect_practical_effect(
    model,
    term = "wt",
    minimum = 2,
    direction = "negative"
  )
)

claims
```

Every result contains a stable claim identifier, status, observed value,
expectation, term, message, evidence, diagnostic details, and originating call.

```{r inspect}
str(claims$results[[1]], max.level = 1)
summary(claims)
```

## Report and export results

`report_claims()` returns a Markdown table suitable for a Quarto or R Markdown
document.

```{r markdown-report, results = "asis"}
cat(report_claims(claims, format = "markdown"), "\n")
```

For further formatting or archival export, convert the suite to a data frame.

```{r export}
claim_data <- as.data.frame(claims)
claim_data

export_path <- tempfile(fileext = ".csv")
utils::write.csv(claim_data, export_path, row.names = FALSE)
file.exists(export_path)
unlink(export_path)
```

The data frame can also be passed directly to `knitr::kable()`.

```{r kable, results = "asis"}
knitr::kable(claim_data)
```

## Make a claim enforceable in CI

Use `expect_claim_passes()` inside a testthat test when a failed scientific claim
must fail continuous integration.

```{r testthat-bridge}
if (requireNamespace("testthat", quietly = TRUE)) {
  expect_claim_passes(
    expect_negative_effect(model, term = "wt")
  )
}
```

For example, a project can place the following code in `tests/testthat/`:

```r
testthat::test_that("the primary conclusion remains valid", {
  model <- run_analysis()
  claimtestR::expect_claim_passes(
    claimtestR::expect_negative_effect(model, term = "treatment")
  )
})
```

## Check stability

The initial stability API accepts existing resample estimates or can refit base R
models with row bootstrap and leave-one-out analyses.

```{r stability}
stability <- expect_stable_direction(
  model,
  term = "wt",
  method = "bootstrap",
  iterations = 50,
  minimum_proportion = 0.9,
  seed = 2026
)
stability
```

Row bootstrap assumes that rows are the appropriate independent sampling units.
Clustered and longitudinal analyses require cluster-aware resampling, which is
planned for a later release and must not be approximated with this method.

## Scope and limitations

`claimtestR` checks claims chosen by the analyst. It does not infer claims from
prose, select a statistical model, determine whether observations are
independent, or establish that a chosen threshold is scientifically justified.
Those decisions remain part of the study design and analysis plan.

